b---
title: "Modelling the annual sentenced prisoner population"
subtitle: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Caleb Moses"
documentclass: scrartcl
# bibliography: ../deepspeech.bib
# selfcitation: ../deepspeech-self.bib
clean: false
output:
  bookdown::pdf_document2:
    toc: true
    latex_engine: pdflatex
    fig_width: 5
    fig_height: 3
    fig_caption: true
    template: arsclassica.tex
    keep_tex: true
---

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center',
  fig.width = 5
  )
```

```{r setup}
library(tidyverse)
library(tidybayes)
library(rethinking)
library(kableExtra)
library(knitr)
library(furrr)
library(rstan)
library(brms)
library(here)
library(gt)

show_priors = FALSE

# Set multicore
options(mc.cores = min(parallel::detectCores() - 1, 10),
        future = FALSE)

plan(multiprocess)

# Set default ggplot theme
relative_font_size = 0.8
theme_set(
    theme_minimal() +
    theme(axis.text.x=element_text(size=rel(relative_font_size)),
          axis.text.y=element_text(size=rel(relative_font_size)),
          axis.title.x=element_text(size=rel(relative_font_size)),
          axis.title.y=element_text(size=rel(relative_font_size)),
          legend.text=element_text(size=rel(relative_font_size)),
          legend.title=element_text(size=rel(relative_font_size)),
          legend.position="bottom")
)

# Avoid recompilation of unchanged models
rstan_options(auto_write = TRUE)

source(here('scripts/zero_inflated_poisson_rr3.R'), local = TRUE)

```

```{r load-data}

fix_factor_levels <- function(x) {
    levels <- levels(x)
    res = unique(as.character(levels))
    end = c()
    if ("Other" %in% levels) {
        res <- res[res != "Other"]
        end <- c("Other")
    }
    if ("Unknown" %in% levels) {
        res <- res[res != "Unknown"]
        end <- c(end, "Unknown")
    }
    c(res, end)
}

duration_levels <- c(
    "Preventive detention",
    "6 months or less",
    "Over 6 months and up to 1 year",
    "Over 1 year and up to 2 years",
    "Over 2 years and up to 3 years",
    "Over 3 years and up to 5 years",
    "5 years or more",
    "Life imprisonment",
    "Unknown")

prison_pop <- read_csv(here('data/processed/prison_pop.csv')) %>%
    left_join(read_csv(here('data/processed/demographics.csv'))) %>%
    left_join(read_csv(here('data/processed/pop_estimates.csv'))) %>%
    mutate_if(is.character, factor) %>%
    mutate_if(is.numeric, as.integer) %>%
    mutate_if(is.factor, function(x) factor(x, levels = fix_factor_levels(x))) %>%
    mutate(Age = as.ordered(Age),
           Duration = ordered(Duration, levels = duration_levels)) %>%
    select(Year, Sex, Age, Ethnicity, Offence, Duration, Prisoner_Count,
           Population_Count, Estimated_Count, Sex_Total, Age_Total,
           Ethnicity_Total, Offence_Total, Duration_Total) %>%
    arrange(Year, Sex, Age, Ethnicity, Offence, Duration) %>%
    distinct()

```

# Introduction

I have collected the Annual Sentenced Prisoner Population for the latest Fiscal Years statistics, released annually by Stats NZ via the NZ.Stat website (http://nzdotstat.stats.govt.nz/).

The data is provided as randomly rounded counts^["Privacy, security, and confidentiality of information supplied to Statistics NZ" available here: https://bit.ly/2A9O22r], and contains the following variables.

- Year
- Sex
- Age
- Ethnicity
- Offence
- Duration
- Sentenced prisoner count
- Population count

A list of the possible values for each variable is provided in table \@ref(tab:varsummary).

Source Tables:

- Annual Sentenced Prisoner Population for the latest Calendar Years (ANZSOC)

- Estimated resident population (ERP), national population by ethnic group, age, and sex, 30 June 1996, 2001, 2006, and 2013

- Age and sex by ethnic group (grouped total responses), for census night population counts, 2006, 2013, and 2018 Censuses (RC, TA, SA2, DHB)

```{r varsummary}

prison_vars <- c("Year", "Sex", "Age", "Ethnicity", "Offence", "Duration", "Prisoner_Count", "Population_Count", "Estimated_Count")

prison_pop %>%
    select(-ends_with("Total")) %>%
    summarise_all(function(x) list(levels(x))) %>%
    t() %>%
    as_tibble() %>%
    rename(Values = V1) %>%
    mutate(Variable = prison_vars) %>%
    unnest(cols = Values, keep_empty = TRUE) %>%
    mutate(Variable = factor(Variable, levels = unique(Variable))) %>%
    mutate(Variable = factor(str_replace_all(Variable, "_", " "),
                             levels = str_replace_all(prison_vars, "_", " ")),
           Values = factor(str_wrap(Values, width = 60))) %>%
    group_by(Variable) %>%
    summarise(Values = paste0(Values, collapse = "\n")) %>%
    mutate(Values = case_when(
               Variable == "Year" ~ sprintf("%d-%d", min(prison_pop$Year), max(prison_pop$Year)),
               Variable == "Prisoner Count" ~ "Count of prisoners, randomly rounded to base 3",
               Variable == "Population Count" ~ "Population counts, randomly rounded to base 3",
               Variable == "Estimated Count" ~ "Estimated New Zealand resident population"
               TRUE ~ Values)) %>%
    mutate(Values = linebreak(Values)) %>%
    rename(`Variable name` = Variable, `List of values` = Values) %>%
    kable(escape = FALSE, booktabs = TRUE, linesep = "\\midrule", caption="An exhaustive list of values in the Annual Sentenced Prisoner Population dataset") %>%
    kable_styling(font_size = 10)

```

# See the data

\begin{landscape}

```{r offences-by-ethnicity, fig.cap=sprintf("Offences by ethnicity from years %d to %d", min(prison_pop$Year), max(prison_pop$Year)), fig.height=6, fig.width=10}

prison_pop %>%
    group_by(Year, Offence, Duration) %>%
    summarise(Prisoner_Count = sum(Prisoner_Count)) %>%
    ungroup() %>%
    filter(Duration != "Unknown", Offence != "Unknown") %>%
    mutate(Offence = str_wrap(Offence, width = 20)) %>%
    ggplot(aes(x = Year, y = Prisoner_Count, colour = Duration)) +
    geom_line(stat = 'identity') +
    facet_wrap(~Offence, scales = 'free_y') +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank())

```

\end{landscape}

# Model the data

```{r offence-duration-formula, fig.height=5.5}

severity_form <- bf(Prisoner_Count ~ 1 + Offence + Duration, decomp = 'QR')

severity_data <- prison_pop %>%
    filter(Year == 2018) %>%
    group_by(Offence, Duration) %>%
    summarise(Prisoner_Count = sum(Prisoner_Count))

if (show_priors) {
    severity_prior <- brm(
        severity_form, data = severity_data,
        family = poisson,
        prior = c(prior(normal(0, 0.9), class = Intercept),
                  prior(normal(0, 0.6), class = b)),
        sample_prior = "only",
        save_model = here('models/severity_fit.stan'),
        file = here('models/severity_prior.rds')
        )

    severity_data %>%
        add_fitted_draws(severity_prior, value = 'Pred_Prisoner_Count') %>%
        select(.draw, Offence, Duration, Prisoner_Count, Pred_Prisoner_Count) %>%
        group_by(.draw, Offence) %>%
        summarise(Prisoner_Count = sum(Prisoner_Count),
                  Pred_Prisoner_Count = sum(Pred_Prisoner_Count)) %>%
        mutate(Offence = str_wrap(Offence, width = 35)) %>%
        ggplot(aes(x = reorder(Offence, Prisoner_Count))) +
        stat_interval(aes(y = Pred_Prisoner_Count)) +
        scale_color_brewer() +
        geom_point(aes(y = Prisoner_Count), shape = 21, size = 1.3, fill = 'white', stroke = 1) +
        coord_flip() +
        xlab("Offence Type") +
        ylab("Annual Sentenced Prisoner Population")
}

```

```{r pred-offence-counts, fig.height=5.5, fig.cap="Simulated Annual Sentenced Prisoner Population counts by offence type. The white dots give the reported (actual) data, while the blue bands report 50\\%, 80\\% and 95\\% confidence intervals."}

severity_fit <- brm(
    severity_form, data = severity_data,
    family = poisson,
    prior = c(prior(normal(0, 0.9), class = Intercept),
              prior(normal(0, 0.6), class = b)),
    save_model = here('models/severity_fit.stan'),
    file = here('models/severity_fit.rds')
    )

severity_data %>%
    add_fitted_draws(severity_fit, value = 'Pred_Prisoner_Count') %>%
    select(.draw, Offence, Duration, Prisoner_Count, Pred_Prisoner_Count) %>%
    group_by(.draw, Offence) %>%
    summarise(Prisoner_Count = sum(Prisoner_Count),
              Pred_Prisoner_Count = sum(Pred_Prisoner_Count)) %>%
    mutate(Offence = str_wrap(Offence, width = 35)) %>%
    ggplot(aes(x = reorder(Offence, Prisoner_Count))) +
    stat_interval(aes(y = Pred_Prisoner_Count)) +
    scale_color_brewer() +
    geom_point(aes(y = Prisoner_Count), shape = 21, size = 1.3, fill = 'white', stroke = 1) +
    coord_flip() +
    xlab("Offence Type") +
    ylab("Annual Sentenced Prisoner Population")

```
